---
title: "Practicals 4"
author: "January Weiner"
date: "5/6/2021"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Session 1: Introduction to R

```{r}
# crap. It did not save it, grr
## (my oldschool editor saves automatically, so I don't think about it)

#### asdff asdf asdff ####

5 + 7
a <- 5 + 7
5 ** 7 # obsolete, but works
5^7 # same as 5 ** 7

## define function blah
#' @param par1 first parameter. Does absolutely nothing
## By the way, this previous line is a formalized way of documenting functions.
## This comes with the roxygen2 package and is used to document packages.
myfunc <- function(par1) {
  
  ## asdfasdf
  
  
}
```

Here, we play with operators.

```{r operators}


## what does the following do?
5 %/% 7
5 / 7
5 %% 7


## to get help on operators, you need to use quotes
?"%*%"
?"%%" # what is this operator
```

So, what do these operators do?

And now for vectorization. Vectorization is really important: everything in R is either a vector, or a more complex data type. Vectors can have length of 1, or even 0 (yes, that is possible)

```{r vectorization}
## everything is a vector!
v <- c(1, 2, 3) 
is.vector(v)
v <- TRUE
is.vector(v)
v <- 1
is.vector(v)

## vectorization, you need to 
c(1, 2, 3) * 5
c(1, 2, 3) * c(2, 5) # error message!
c(1, 2, 3) * c(1, 5, 10)
## scalar product
c(1, 2, 3) %*% c(1, 5, 10)
```



## First statistical tests

We define two variables:

$$x \sim f_N(\mu, \sigma)$$

with $\mu = 0$ and $\sigma = 0$, as well as

$$y = x + \epsilon$$

where

$$\epsilon \sim f_N(\mu, \sigma)$$
And now, for some plotting. The `fig.height` and `fig.width` parameters for the chunk determine the width and height of the resulting figure.

```{r plotting,fig.width=7,fig.height=7}
## get random numbers and plot them
x <- rnorm(100)
y <- x + rnorm(100)
plot(x, y)
```

```{r correlations}
veryveryververyvevyresdrcc <- function()  { }
## 
cor(x,y)
cor(x,y,method="s")
cor.test(x,y)
cor.test(x, y, method = "s")
```


```{r}
myfunc <- function(param1) {
  
  param1 <- 10
  x <- rnorm(100)
  y <- x + rnorm(100)
  ret <- list(xx=x, yy=y)
  return(ret)
}
```

