---
title: "Writing the report"
subtitle: "BE_22 Bioinformatics SS 21"
author: "January Weiner"
date: "`r Sys.Date()`"
outputdupa: html_document
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: true
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
library(pander)
library(knitr)
library(kableExtra)
source("functions.R")
```

```{r echo=FALSE,include=FALSE,eval=TRUE}
options(crayon.enabled = TRUE)
options(crayon.colors = 256)
knitr::knit_hooks$set(output = function(x, options){
  paste0(
    '<pre class="r-output"><code>',
    fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
    '</code></pre>'
  )
})

## this is an ugly, ugly hack, but otherwise crayon does not LISTEN TO REASON!!!
num_colors <- function(forget=TRUE) 256
library(crayon)
assignInNamespace("num_colors", num_colors, pos="package:crayon")
```


```{r libraries,cache=FALSE}
library(tidyverse)
```




## Introduction to the report

The goal of the report is to demonstrate your ability to write a
scientific, reproducible report.

You should not spend too much time on it; also, submitting a half-baked,
unfinished, error-ridden report is **infinitely better** than not submitting
anything at all.

My estimated effort of writing the entire report is 2-3 hours if you work
quickly (and were present during the practicals), and a day if not.

## rules ✪

 * *In any case*, you will do the report in Rmarkdown or Jupyter lab book.
 * There are three main options:
  * Differential gene expression analysis, combined with some functional
     analysis
  * Writing your own Python or R program (some ideas will follow) that actually
    does calculate something on its own
  * Anything else that you have came up with
 * The main criterion is that your program or your Jupyter notebook or your
   Rmarkdown actually works and can produce a document (HTML, Word,
   whatever)
 * Report is 30% of the grade! (but you can get more, see below!)


I encourage you to use whatever help which is necessary; you can use
Codepilot, Wikipedia, ChatGTP... however: 

 * no plagiarism. Getting caught with plagiarism is automatic 0 points.
 * if you blindly copy code from somewhere, without understanding
   what it does, it probably will not work. Make sure that the code works,
   and make sure that you understand what it does. 
 * Should any doubts about your code or the authorship of your report
   arise, you might be asked to explain your code or your report.
   
 

## Report on differential gene expression analysis

 * Run a basic differential gene expression analysis as shown in the
   course
 * There will be three different data sets, each consisting of a matrix of
   count data. Make sure you write which of the data sets you are using.
 * Look up the data set in the GEO database and write a short description
   of the experiment in the "Introduction" part.
 * Create a visualisation of the results (volcano plot)
 * In addition, do one of the following:
   * Select one of the top results and describe its structure *or*
   * Find homologs and make an MSA + phylogenetic tree, *or*
   * Run a gene set enrichment analysis
 * Document all things you have done in a *replicable* manner
 * Use the "Discussion" part to describe your opinion on the results,
   speculations, and what els could one do



## Report on your own program

 * document the program in your Rmarkdown document / Jupyter lab book (what
   does it do? What did you do with it? what were the results? what are the
   limitations?)
 * Document all things you have done in a *replicable* manner
 * Use the Discussion part to show what else could one do with the program,
   or how it could be expanded
 * It is more important that the program works and that you have documented
   it well than that is has all the features one could dream of



## Writing the report

 * Once you have completed your tasks, write the report
 * Report must be written in Rmarkdown or Jupyter lab book
 * You must create a bibliography file and include the references
 * Like a scientific paper, the report must include the following parts:
    * Abstract: short description of your work
    * Introduction: explain what is the background of your work (e.g. why
      analysing proteins is important)
    * Methods: describe, *in detail*, the methods you will apply. Cite sources!
    * Results: describe, *with figures*, your findings. 
    * Discussion: what is your opinion? what else should / could one do?  What did you learn?
    * Bibliography


There are three types of statements:

 * Statements by others: you must cite them.
 * Statements derived from your findings: you must back them up with results
 * Statements which are speculative: you must label them clearly as such


## Marking criteria

 * Structure of the report
 * Reproducibility of the results (i.e., I can run the markdown / jupyter
   notebook)
 * Citing the sources
 * Documenting the code
 * There is a bonus (generally 5% of the grade) for 
  - doing something extraordinary
  - making me laugh
  - doing something really smart
  - coming up with a really good idea 
 * The bonus allows you to get 100% of the final mark, even if you have not
   gotten full marks on the exam or the homeworks.
 * What is **not** important for the marking:
   - whether the code actually is correct (as long as it runs!)
   - whether you copied the code from practicals or the internet
   - whether your statements are correct
   - whether your results are correct
   - whether your program is useful
   - whether your program is well written
   - whether your report is long and full of complicated words 

## Submission

You will submit a single `.zip` or `.tar.gz` or `.rar` file. The file name
must include your first name, last name, immatriculation number and the
word "report" (e.g. `january_weiner_1234567_report.zip`).
The archive should contain the following:

 * The Rmarkdown / Jupyter notebook file
 * The generated HTML / Word / PDF report
 * Any further data necessary to run the report (e.g. the data files)

Please make sure that the report source code does not contain absolute 

**Submission date:** 2021-07-14 23:59:59 CEST
