---
title: "Practicals 05: Sequence reconstruction with Phylip"
subtitle: "BE_22 Bioinformatics SS 21"
author: "January Weiner"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: true
      toc_float: true
outputxxx:
  xaringan::moon_reader:
    self-contained: true
    css: ["default", "files/cubi-fonts.css", "files/style.css" ]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "files/macros.js"
toc: no
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE}
```

### Installation

First, we need to install a few packages. We will also install BiocManager
(Bioconductor package manager), because we will need it during following
practicals.

```{r, eval=FALSE}
install.packages(c("phangorn", "seqinr", "BiocManager"))
BiocManager::install("msa")
```

This will take a moment, during which we will talk about what the plan for
the practicals is.

### Loading the multiple sequence alignment

It is possible, of course, to create multiple sequence alignments using R
(e.g. with the `msa` package from Bioconductor). However, it is also
possible to read in an MSA in the phangorn package.
Here, we will start with an amino acid alignment. 

As you can see in the script that follows, you can supply an URL to R and
it will work just like a file. However, normally you could just use a
regular file name.

We will start with our beloved hemoglobin example.

```{r reading_msa}
library(phangorn)
library(seqinr)
url <- "https://january3.github.io/Bioinformatics/Practicals/practicals_02_220421/data/hba.aln"
msa <- read.alignment(url, format="clustal")
```

The `msa` object has been created by the seqinr package and we need to
modify it a bit before proceeding.

```{r modify_msa}
## remove the boring identifiers from the sequence names,
## so "sp|P01942|HBA_MOUSE" becomes "MOUSE". Don't do it unless all
## proteins are HBA_something or HBA2_something
msa$nam <- gsub(".*(HBA_|HBA2_)", "", msa$nam)

## convert to phangorn package object
msa <- as.phyDat(msa, type="AA")
```

The `msa` object holds now the multiple sequence alignment. We had to
convert it to a data typ that is suitable for the `phangorn` package,
though.

We will now use
the `dist.ml` function from the `phangorn` package to calculate pairwise
distances between the sequences. We will use the Blosum62 matrix to
calculate the distances:

```{r, dist_mat,results="markdown"}
dist_mat <- dist.ml(msa, model="Blosum62")
dist_mat
```

**Note:** For a more advanced analysis, one would first spent quite some
time deciding which model to chose. This is especially important for DNA
sequences.

We can make a plot visualizing this distance matrix, first converting it to
regular square matrix, and then plotting a heatmap:

```{r heatmap}
mtx <- as.matrix(dist_mat)
heatmap(mtx, symm=TRUE)
```

---

**Exercise 1.**

 1. What does `symm=TRUE` do? (Hint: `?heatmap`)
 2. Which two species seem to be closest? Take a look at the `mtx` and
    check which distances is smallest.
 3. What is BALAC? (hint: the protein name is `HBA_BALAC`...)

---

### Reconstructing phylogenies

We now can do the first phylogenetic reconstruction using neighbor joining.


```{r nj_tree,results="markdown"}
nj_tree <- NJ(dist_mat)
```

That wasn't hard. Now, we plot the tree. 

```{r plot_nj,fig.width=10}
## split the window in two
par(mfrow=c(1,2))

plot(nj_tree)
plot(nj_tree, type="unrooted")
```

Let us see the tree in Newick format as well:

```{r newick_nj_0,results="markdown"}
write.tree(nj_tree)
```

---

**Exercise 2.**

Look up the help for the function `plot.phylo` which is called when you
plot a tree. What are the other types of trees? View them all.

---

The tree on the left sucks: clearly, GADMO (Gadus morhua, atlantic cod) and
POGSC (Saddleback, also a fish) are the outgroup here. We need to *root* the tree
correctly.

```{r root_nj}
nj_tree <- root(nj_tree, outgroup=c("GADMO", "POGSC"))

plot(nj_tree)
```

---

**Exercise 3.**

Instead of the function `NJ`, use the function `upgma()` to create a tree.
Compare it with the neighbor-joining tree. What do you notice?

---

### Bootstrapping a tree

Bootstrapping is an important technique to understand which parts of the
tree are robust, and which are somewhat shaky. Basically, 100 or 1000 trees
are computed, but each time a random subset of the alignment columns is
used. If a part of a tree is solid, this will have no influence; but
othertimes, selecting a subset of columns will change the phylogeny.

Generating bootstrap values may look like that:

```{r bootstrap_nj}
## set up the function to calculate phylogenies
make_phylo <- function(msa) {
  tree <- NJ(dist.ml(msa, model="Blosum62"))
}

boot <- bootstrap.phyDat(msa, make_phylo)
```

The `boot` object contains now a 100 of different phylogenetic trees. We
can plot some of them.

```{r plot_bs_trees,fig.width=15,fig.height=10}
par(mfrow=c(2,3))
for(i in 1:6) {
  plot(boot[[i]])
}
```

Can you see the differences?

We can now plot the bootstrap values on the original tree. The numbers
indicate how often, among the 100 bootstrap replicates, a given group was
found. We use the function plotBS to generate a tree which contains
bootstrap values as node labels:

```{r plot_nj_tree,fig.width=15}
nj_tree_bs <- plotBS(nj_tree, boot, type="none")
plot(nj_tree_bs, show.node.label=TRUE)
```

A slightly more readable figure can be achieved as follows:

```{r plot_nj_tree_2,fig.width=15}
plot(nj_tree_bs)
nodelabels(nj_tree_bs$node.label)
```

Take a look on how the tree looks like in the Newick format:


```{r newick_nj}
write.tree(nj_tree_bs)
```


### Using maximum likelihood for phylogeny reconstruction

As noted, ML methods do not actually reconstruct a tree *per se*. Rather,
they are able to provide a measure called *likelihood* which allows us to
decide which tree is better.

We can now calculate the likelihood function and the AIC (Akaike
Information Criterion) on the NJ tree we have constructed. Note that we
must use the `msa_phydat` object:

```{r calc_nj_ml,results="markdown"}
nj_tree_pml <- pml(nj_tree, msa, model="Blosum62")
nj_tree_pml
AIC(nj_tree_pml)
```

Note that `nj_tree_pml` is no longer an object of class `phylo`: pml added
extra information to that tree. The actual tree is stored in
`nj_tree_pml$tree`. Yeah, I know it is confusing. Welcome to R and bioinformatics!

However, this is still our boring NJ-generated tree. We can try to find a
better one, using some heuristics which starts with an initial tree and
then tries to improve it:

```{r ml_tree, fig.width=12}
ml_tree <- optim.pml(nj_tree_pml, rearrangement="NNI")
ml_tree
AIC(ml_tree)

par(mfrow=c(1,2))
plot(nj_tree)
plot(ml_tree)
```

What differences do you see?

---

**Exercise 4.**

 1. There are various models (both for nucleic acids and for protein
sequences). For example, for amino acid sequences there are the models
Blosum62, Jones-Taylor-Thornton (JTT), Dayhoff (of the PAM matrix fame),
Whelan and Goldman (WAG).  Try all of them; which of them fits the data
best? (the higher the log likelihood, the better; the lower the AIC, the
better).

 2. Plot all the trees you have created in the previous point (WAG, JTT,
    Dayhoff, Blosum62) and compare them to the NJ tree.

---


### Bootstrapping the ML tree (optional)

We are going to use a bootstrapping function from the `phangorn` package.
The plotBS function not only plots the tree, it also returns a tree with
the node labels.

```{r ml_bootstrap,fig.width=15}
boot_ml <- bootstrap.pml(ml_tree)
boot_tree <- plotBS(ml_tree$tree, boot_ml, type="none")
plot(boot_tree)
nodelabels(boot_tree$node.label)
```

As you can see, ML is pretty sure about its result – all bootstrap values
are 100! However, it is also wrong, since horse should be in one group with
cattle. 

**Homework 4.**

Use the data from 
`https://january3.github.io/Bioinformatics/Practicals/practicals_02_220421/data/globin.aln`
to reconstruct phylogeny using a suitable amino acid substitution model. As
outgroups, use the sequence of myoglobin (`MYG_HUMAN`).  Just like in case
of homework 3, you are expected to provide an Rmd file which can be
compiled without errors.
