---
title: "Practicals 08: Differential expression and gene set enrichments"
subtitle: "BE_22 Bioinformatics SS 21"
author: "January Weiner"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: true
      toc_float: true
outputxxx:
  xaringan::moon_reader:
    self-contained: true
    css: ["default", "files/cubi-fonts.css", "files/style.css" ]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "files/macros.js"
toc: no
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE,fig.width=7,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
options(crayon.enabled = TRUE)
options(crayon.colors = 256)
knitr::knit_hooks$set(output = function(x, options){
  paste0(
    '<pre class="r-output"><code>',
    fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
    '</code></pre>'
  )
})

## this is an ugly, ugly hack, but otherwise crayon does not LISTEN TO REASON!!!
num_colors <- function(forget=TRUE) 256
library(crayon)
assignInNamespace("num_colors", num_colors, pos="package:crayon")
```

```{r libraries,cache=FALSE,echo=FALSE}
library(ggplot2)
library(tidyverse)
library(DESeq2)
library(tmod)
library(edgeR)
theme_set(theme_bw())
options(colorDF_n=30)
options(width=100)
```

```{r eval=FALSE,echo=FALSE}
## this is how the counts45 data was derived

annot <- readRDS("../../Datasets/annot.rds")
covar <- readRDS("../../Datasets/covar.rds")
covar45 <- covar %>% group_by(group) %>% sample_n(15)
counts <- readRDS("../../Datasets/counts.rds")
all(colnames(counts) %in% covar$label) ## TRUE
counts45 <- counts[ , covar45$label ]
library(edgeR)
vars <- apply(cpm(counts45, log=TRUE), 1, var)
sel <- vars > quantile(vars, .25)
all(rownames(counts) == annot$PrimaryID) ## TRUE
counts45 <- counts45[sel, ]
annot45  <- annot[sel, ]
annot45 <- annot45 %>% dplyr::select("ENSEMBL", "SYMBOL", "GENENAME")
all(rownames(counts45) == annot45$PrimaryID) ## TRUE
all(colnames(counts45) == covar45$label) ## TRUE

saveRDS(counts45, file="../../Datasets/counts45.rds")
saveRDS(annot45, file="../../Datasets/annot45.rds")
saveRDS(covar45, file="../../Datasets/covar45.rds")
```


### Preparation

Download the files `counts45.rds`, `covar45.rds` and `annot45.rds` from the
github repository. These RDS files contain, respectively, the counts,
the covariates and gene annotation for 45 samples from the data set 
[GSE156063](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE156063).
These RDS files contain R objects saved with the `saveRDS` function. 

The only reason we use the RDS files is that they are smaller than the
corresponding TSV/CSV text files. You can load them as follows:

```{r,eval=TRUE,echo=FALSE}
counts <- readRDS("../../Datasets/counts45.rds")
covar  <- readRDS("../../Datasets/covar45.rds")
annot  <- readRDS("../../Datasets/annot45.rds")
```


```{r,eval=FALSE}
counts <- readRDS("counts45.rds")
covar  <- readRDS("covar45.rds")
annot  <- readRDS("annot45.rds")
```

There are three groups of data in the covariate file.

```{r results="markdown"}
library(colorDF)
summary_colorDF(covar)
```

**Exercise. (10 minutes)** Take a look at the `covar` object. How many females and how
many males are there? How many females have COVID-19, and how many males
are healthy? What is the average age in the male and female groups? Hint:
use the `table` function.


## Differential gene expression analysis

To run a basic gene expression analysis, we need to use the DESeq2 package.
The steps are as follows:

 * Create the DESeq2 object with counts and covariates
 * Run the calculations – fit the model
 * Extract the results

First two steps are below:

```{r, cache=TRUE}
library(DESeq2)
ds2 <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=covar,
                              design=~ group)
ds2 <- DESeq(ds2)
```

The ds2 contains now the fitted model. We now have to get the results, but
for that, we need to know the names of the results:

```{r cache=TRUE,results="markdown"}
resultsNames(ds2)
res_sc2 <- results(ds2, name="group_SC2_vs_no")
res_sc2 <- as.data.frame(res_sc2)
print_colorDF(head(res_sc2))
```

This is not very interesting, unless we merge it with the annotation table
and order by p-values. The row names of `res_sc2` correspond to the column
"ENSEMBL" of `annot`:

```{r results="markdown"}
## merge by ENSEMBL column from annot and rownames ("0")
## from res
res_sc2 <- merge(annot, res_sc2, by.x="ENSEMBL", by.y=0) %>%
  arrange(pvalue)
knitr::kable(head(res_sc2))
```

**Exercise. (5 minutes)**. How many genes are significant at padj < 0.05?
How many at padj < 0.01? (padj is the false discovery rate from the
Benjamini-Hochberg procedure). How many are significant at padj < 0.05 and
abs(log2FoldChange) > 1?

**Exercise. (15 minutes)**. Repeat the above steps for the other comparison
(other virus vs. no virus) generating a suitable data frame. Merge the two
result tables and plot the log2 fold changes from one comparison vs the
log2 fold changes from the other comparison. What do you notice?

**Exercise. (extra)** The model `~ group` fits the data using `group` as predictor.
To find genes that are related to age, you would use a model like `~ age`.
Using DESeq2, find genes that show a significant association with age
(repeat the steps above with a new model).


## Visualizing the results

To visualize the results, we need to first calculate the log-counts per
million. One could use here the `cpm` function from the package `edgeR`,
but we will do it manually. We need, however, to add a small value to all
results to avoid log2(0).

```{r}
cpm <- counts + 2
lib_sizes <- colSums(cpm) / 1e6 # library sizes in millions
## a neat trick to divide a matrix by a row
cpm <- t(t(cpm) / lib_sizes)
cpm <- log2(cpm)
```


Now we can plot the top results.

```{r}
par(mfrow=c(1,2))
id <- "ENSG00000126709" # IFI6
boxplot(counts[id, ] ~ covar$group)
boxplot(cpm[id, ] ~ covar$group)
```

**Exercise. (5min)** Choose another gene and plot it.


To create a heatmap, we can use the very simple `heatmap` function. We will
select the top 25 genes for that:

```{r}
sel <- res_sc2$ENSEMBL[1:25]
x <- cpm[sel, ]
colnames(x) <- covar$group
rownames(x) <- annot$SYMBOL[ match(sel, annot$ENSEMBL) ]
heatmap(x)
```

**Exercise. (15min)** Create PCA based on the CPM of the top 500 genes (see
practicals 5 and 7).


## Gene set enrichment analysis

### Gene set enrichment

Gene set enrichment analysis with tmod is as simple as this:

```{r results="markdown"}
library(tmod)
tmodCERNOtest(res_sc2$SYMBOL)
```

**Exercise. (20min)** Run the enrichment analysis for the other comparison.
Compare the results of the enrichment (AUC, p-values) on a plot.

### ORA (Overrepresentation Analysis)

We will select genes from a single gene set and test them using a $\chi^2$
test.

```{r}
degs <- abs(res_sc2$log2FoldChange) > 1 & res_sc2$padj < 0.05
degs <- ifelse(degs, "DEG", "NS")
data(tmod)
ifn <- tmod$MODULES2GENES[["LI.M75"]]
ings <- res_sc2$SYMBOL %in% ifn
ings <- ifelse(ings, "IFN", "Other")
table(ings, degs)
```

```{r}
chisq.test(table(ings, degs))
```

## Other visualizations

### Evidence plot

```{r}
evidencePlot(res_sc2$SYMBOL, m="LI.M75", gene.labels = TRUE)
```

## Panel plot

This one is a bit more complex.

```{r}

res <- list()
res$sc2 <- results(ds2, name="group_SC2_vs_no")
res$sc2 <- as.data.frame(res$sc2)
res$other <- results(ds2, name="group_other_vs_no")
res$other <- as.data.frame(res$other)

res$sc2 <- merge(annot, res$sc2, by.x="ENSEMBL", by.y=0) %>%
  arrange(pvalue)
res$other <- merge(annot, res$other, by.x="ENSEMBL", by.y=0) %>%
  arrange(pvalue)

tmod_res <- lapply(res, function(x) tmodCERNOtest(x$SYMBOL)

```


